---
title: Why does the cost to save a life differ dramatically by country? An empirical
  test of the power-weighted social decision rule
output:
  pdf_document:
    fig_caption: yes
bibliography: cost_effectiveness.json
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(countrycode)
library(corrplot)
library(dplyr)
library(fredr)
library(fuzzyjoin)
library(ggplot2)
library(lubridate)
library(pander)
library(readr)
library(rnaturalearth)
library(sf)
library(stringi)
library(tidyr)
library(WDI)

panderOptions("digits", 2)

# get the codes for the variables
Public = WDIsearch("Domestic general government health expenditure per capita \\(current US\\$\\)")[["indicator"]]
Private = WDIsearch("Domestic private health expenditure per capita \\(current US\\$\\)")[["indicator"]]
`Non-Profit` = WDIsearch("External health expenditure per capita \\(current US\\$\\)")[["indicator"]]

CPI_data = 
  fredr(
    # CPI
    series_id = "CPIAUCSL",
    observation_start = as.Date("2010-01-01")
  ) %>%
  # average by year
  mutate(year = year(date)) %>%
  group_by(year) %>%
  summarize(value = mean(value))

# to a named vector
CPI = CPI_data$value
names(CPI) = CPI_data$year

expenditures = 
  WDI(indicator = c(
    Public,
    Private,
    `Non-Profit`
  )) %>%
  rename(
    # interpolate in codes for the variables
    Public = {{Public}},
    Private = {{Private}},
    `Non-Profit` = {{`Non-Profit`}},
    country_code = iso2c
  ) %>%
  # we only need the contry code
  select(-country)

fuzzy_countries = 
  codelist %>%
  select(
    country = country.name.en,
    country_code = iso2c
  ) %>%
  # we want Taiwan to be part of China to match the World Bank (politics...)
  filter(country != "Taiwan") %>%
  mutate(
    # add word boundaries
    fuzzy_country = paste0("\\b", country, "\\b")
  ) %>%
  select(-country) %>%
  # add some alternate names, demonyms, and misspellings
  bind_rows(read_csv("fuzzy_country.csv"))

ratios_duplicated = 
  read_csv("cost_effectiveness.csv") %>%
  select(
    year = `Publication Year`,
    fuzzy_country = `Target Population`,
    `2018 $/QALY` = `US$/QALY *`
  ) %>%
  # remove results duplicated in searches
  distinct %>%
  # assign an ID
  mutate(ratio_id = 1:n()) %>%
  regex_left_join(fuzzy_countries) %>%
  select(-fuzzy_country.x, -fuzzy_country.y)

ratios = 
  ratios_duplicated %>%
  # count
  group_by(ratio_id) %>%
  summarize(count = n()) %>%
  filter(count == 1) %>%
  # keep if exactly one country matches
  semi_join(ratios_duplicated, .) %>%
  select(-ratio_id) %>%
  mutate(
    `VEL (2010 $/70 QALYs)` = as.numeric(ifelse(
        # ignore non-numeric ratios
        # \u00A0 = no break space
        `2018 $/QALY` %in% c("Cost-Saving", "Dominated", " ", "\u00A0"),
        NA, `2018 $/QALY`
      )) * CPI["2010"] / CPI["2018"] * 70
  ) %>%
  filter(!is.na(country_code) & !is.na(`VEL (2010 $/70 QALYs)`)) %>%
  group_by(year, country_code) %>%
  summarize(`Mean log VEL (2010 $/70 QALYs)` = mean(log(`VEL (2010 $/70 QALYs)`))) %>%
  ungroup %>%
  # match to world bank data two years before
  mutate(year = year - 2)

together = 
  ratios %>%
  left_join(expenditures) %>%
  mutate(
    `Log health care spending per capita (2010 $)` = 
      log(
        (Public + Private + `Non-Profit`) * 
        CPI["2010"] / 
        CPI[as.character(year(Sys.Date()))]
      )
  )

model = 
  lm(
    `Mean log VEL (2010 $/70 QALYs)` ~ `Log health care spending per capita (2010 $)`,
    together
  )

interval = confint(model)["`Log health care spending per capita (2010 $)`",]
```

# Abstract

I attempted to empirically test the power-weighted social decision rule, which predicts that international non-profits, international organizations, and foreign aid agencies will effectively value the powerful more. I used the cost to save a life by country and year to measure the value of an empirical life (VEL). To accomplish this, I pulled data from the Cost Effectiveness Analysis Registry at Tufts Medical Center. I matched this data to the World Development Indicators from the World Bank. I found VELs to be strongly determined by health care spending, with decreasing returns to spending per capita. I further decomposed variance in health care spending per capita into its private, public, and non-profit components. Differences in public health care spending per capita can explain the majority of this variation. The mere existence of very cost effective ways to save lives in many countries suggests that citizens in these countries do not have the power to effectively influence global health care spending. This highlights the hypocrisy of any claim that "all lives matter".

# Introduction

Cost-effectiveness analysis (CEA) was designed to avoid questions about the value of a human life, but this value has been subtly reintroduced via cost effectiveness thresholds. The history of cost-effectiveness analysis began with studies of the value of a statistical life.

As documented by Banzhaf, the earliest considerations of the value of a statistical life came from air force consultants at RAND. These consultants were trying to decide how much money to spend on cheap yet dangerous airplanes, and how much money to spend on expensive yet safe airplanes. Because their value of a statistical life was implicitly zero, the model recommended building the cheap and dangerous planes [-@Banzhaf_2014, p. 215]. RAND concluded that further progress would require placing a number on the value of human life. A PhD student who formerly worked as a RAND consultant brought the idea to Schelling, who argued: "where life and death are concerned, we are all consumers. We nearly all want our lives extended and are probably willing to pay for it"
[-@Schelling_1968, p. 129].

From the earliest studies, the value of a statistical life was tied to willingness to pay, and thus income. For this reason, in an infamous memo, Summers concluded "the economic logic behind dumping a load of toxic waste in the lowest wage country is impeccable" [-@Summers_1991]. Even the 9-11 compensation fund was required by statute to be awarded based in part on foregone income. The lawyer who oversaw the compensation, Kenneth Feinberg, was not content with this outcome: "any statutory compensation formula that recognizes and reinforces economic distinctions among claimants is guaranteed to promote divisiveness among the very families it seeks to help" [-@Feinberg_2005, p. 183]. The EPA prescribes a constant value for any life. However, this was not done out explicitly out of a concern for equity, but only as a way to reduce informational demands: "An ideal value estimate for fatal risk reduction would account for all of these demographic and risk characteristics...Unfortunately, it is simply too expensive and time-consuming to obtain such unique risk value estimates for each EPA policy" [@EPA_2000, p. 88].

The mere existence of workmen's compensation suggests that workers are not sufficiently compensated for risk via wages. Some researchers have attempted to pull the value of a statistic life from compensating wage differentials, that is, the wage premium paid for dangerous jobs. However, as Dorman has documented, if this differential fully compensated workers for workplace injury, the existing institution of worker's compensation would be obsolete. "If workers are fully compensated for bearing risks *before* they are hurt, any additional compensation afterwards would be both unjust and inefficient" [@Dorman_1996, p. 109]. Thus, wage differentials cannot fully quantify the value of a human life.

The debate about the value of a statistical life came to a head when evaluating global warming damages. Researchers wanted to justify global warming by discounting away the lives of future generations, and justified doing this by weighing benefits to comparatively rich future people lower than benefits to comparatively poor present people. Admitting the decreasing marginal utility of money, however, immediately led to a policy recommendation of global wealth redistribution. To avoid this "unsatisfactory" result, these researches put Negishi weights on richer countries, effectively valuing lives in richer countries more. As Stanton summarizes: "in Negishi-weighted models, human welfare is more valuable in richer regions and redistribution of income among world regions is suppressed, maintaining the existing income distribution" [@Stanton_2010, p. 423].

A key benefit of cost-effectiveness analysis is that you do not need to put a dollar figure on the value of life. Smith seems to have been the first to clarify the framework of cost-effectiveness analysis as opposed to cost-benefit analysis. As Smith explains, "until we discover a universal unit for quantifying the value of health, ways to measure the benefits will continue to hamper comparisons among alternate health programs" [@Smith_1968, p. 900]. This conveniently sidesteps arguments about the incommensurability of life and money. Such arguments, as given by Raz, are "if people\'s evaluations can form a foundation for a general theory of comparative value judgments, perhaps people\'s refusal to evaluate could provide a pointer to the existence of significant incommensurabilities" [-@Raz_1986, p. 129].

Cost effectiveness analysis picked up steam in the 90's, as explained by Neumann: "By the early 1990's, the rising cost of healthcare and the need to prioritize resource allocation more rationally were firmly on the national agenda... In 1993, the US Public Health Service convened a Panel of 13 non-government scientists and scholars\...to provide
recommendations for the use and conduct of CEAs" [-@Neumann_2017, p. xx]. Literature on cost-effectiveness analysis generally falls into one of three categories:

1.  Reviews of a large-number of cost-effectiveness analyses, generally concluding that many public health interventions are cost-effective. For example, in the United Kingdom, Owen et al. found "nearly two-thirds (63%) of public health interventions assessed were cost-effective" [-@Owen_2018, p. 557].
2.  Guidance on the practice of cost-effectiveness analysis. For example, Hutebessy et al. "set out to overcome a number of barriers to the appropriate use of cost-effectiveness" [-@Hutubessy_2003, p. 1].
3.  Applications of cost-effectiveness analysis for specific illnesses or treatments, e.g. exercise programs [@Hatziandreu_1988].

Cost effectiveness thresholds subtly reintroduce questions of the value of human life back into cost effectiveness analysis. This is a threshold that national health services might use to decide which treatments to fund and which treatments not to. These thresholds vary widely in practice, as countries differ significantly in their ability to fund health improvements. At most, this should be an unsettling reality. However, Garber and Phelps argue that under a notion of Pareto efficiency, reality is optimal: "our estimates imply that\...CE cutoffs should be about double the annual income" [-@Garber_1997, p. 25]. This recommendation was endorsed by the World Bank Commission on Macroeconomics and Health: "This multiple of earnings reflects the value of leisure time in addition to market consumption, the pure longevity effect, and the pain and suffering associated with disease" [-@Sachs_2001, p. 31], and then by the WHO: "interventions that avert one DALY for less than average per capita income for a given country or region are considered very cost-effective; interventions that cost less than three times average per capita income per DALY averted are still considered cost-effective; and those that exceed this level are considered not cost-effective" [@Hutubessy_2003, p. 31]. Shillcutt et al. express some ethical misgivings: that this "puts a value on livelihood rather than life" [-@Shillcutt_2009, p. 909]. Nevertheless, it seems that many authors think that these implied thresholds are in fact too high: "the WHO-CHOICE thresholds set
such a low bar for cost-effectiveness that very few interventions with evidence of efficacy can be ruled out" [@Marseille_2015, p. 118].

Effective altruism, an outcome of cost effectiveness analysis, regrettably focuses on individual, not social, obligations. Decades ago, Peter Singer noted: "if it is in our power to prevent something bad from happening, without thereby sacrificing anything of comparable moral importance, we ought, morally, to do it.... If this principle were acted upon, even in its qualified form, our lives, our society, and our world would be fundamentally changed" [-@Singer_1972, p. 231]. Based on this principle, a non-profit organization called GiveWell seeks out and recommends the most cost-effective charities [-@GiveWell].

The power-weighted social decision rule predicts that people with influence will value the powerful more [@Boyce_2002]. In a key empirical test of this rule, Torras and Boyce show that environmental damages are related not to stage of development, but to power: "literacy, political rights, and civil liberties are found to have particularly strong effects on environmental quality in low-income countries" [@Torras_1998, p. 147]. The entire of body of literature on environmental justice, beginning with a United Church of Christ study, gives evidence for this: environmental harms fall disproportionately on poor people and people of color. "Although socio-economic status appeared to play an important role in the location of hazardous waste facilities, race proved to be more significant" [@UCC_1987, p. xiii]. On the international level is the pollution haven hypothesis, which predicts that corporations will move dirty industries to countries with a lower ability to enact environmental protections. Cole suggests "the North may therefore increasingly specialize in 'clean' production and rely on the South for the provision of pollution intensive output" [@Cole_2004, p. 71].

I attempted to use cost-effectiveness analysis data to directly test the predictions of the power-weighted social decision rule.

A couple of examples might be helpful to understand the data intuitively. A quality adjusted life year (QALY) is a health improvement judged as valuable as an additional year of life for someone who is healthy. Cost-effectiveness ratios are reported in 2018 US$ per QALY. As an example, preventing infant diarrhea would have a very high cost-effectiveness ratio: prevention is not only very inexpensive, but would save many years of life as well. However, cancer treatment for someone who is very elderly would have a very low cost-effectiveness ratio: treatment is not only very expensive, but would save very few years of life as well.

If international organizations, non-profits, and foreign aid agencies made spending decisions based on cost-effectiveness, the cost to save a life in a country would reveal exactly how much value a donor on the margin places on a life in that country. I will term this value the value of an empirical life (VEL). To determine how much value donors place on a life, I imagine a market for international aid. The value of a QALY to donors might differ depending on what country a person is from. In an imagined market, by the theory of revealed preferences, the cost to save a life in a country would reveal exactly how much value a marginal donor places on a life in that country. To be clear, I am not arguing that health care spending decisions are made in a market. Instead, I am using an imagined market to motivate the VEL measure calculated below.

The power-weighted social decision rule predicts that, because of their power, international non-profits and foreign aid agencies will effectively value the powerful more [@Boyce_2002]. Thus, the power-weighted social decision rule would predict that the value of an empirical life should be not be the same for everyone. I endeavored to test this hypothesis empirically.

# Calculations

To determine the value of an empirical life by country and year, I used the Cost Effectiveness Analysis Registry [@CEVR_2019]. This registry, maintained by Tufts Medical Center, holds the results from 8,056 public health cost-effectiveness analyses. From these studies, reviewers enter cost-effectiveness ratios, in 2018 $ per QALY. In this data-set, the only demographic variables consistently associated with the cost-effectiveness analysis studies are country and publication year.

I obtained data by searching for the name of every country as listed in the CRAN countrycode package and collecting the first 100 results for each country (the maximum number available without a license). Fortunately, there was only a single country for which the number of results was greater than 100, the United States, and there's no reason to believe restricting data to the 100 most recent entries would bias results. I adjusted the ratios to 2010 \$ per QALY using the US Consumer Price Index. Then, I multiplied by 70 (a typical lifespan). Note that while countries do differ in average lifespans, using a constant lifespan in calculating VELs is necessary to value a year of life equally for everyone. Finally, I took the mean of the log of the VELs by country and year. 

Publication bias might have influenced the results, but it is unlikely that publication bias alone could explain the results. It seems likely that studies which analyze under-used yet very cost-effective treatments are more likely to be published due to their important implications. For publication bias to affect the results here, it would have to have a larger effect in some countries than others. Unfortunately, because no standard error was reported for these estimates, I cannot test for publication bias. However, other meta-reviews suggest that the effect of publication bias on studies of cost-effectiveness does differ by country [@Bell_2006]. Nevertheless, publication bias seems be unable to fully explain the extreme magnitude of the results here.

I matched cost effectiveness data to the World Development Indicators from the World Bank [@WDI] by country and year. Because much of this data is released approximately two years after the fact, I matched VELs to the World Bank data two years previous.

I assumed "domestic private health expenditure per capita (current US\$)" corresponded to private health care spending, "domestic general government health expenditure per capita (current US\$)" corresponded to public health care spending, and "external health expenditure per capita (current US\$)" corresponded to spending from non-profits, international organizations, and foreign aid agencies. Again, I used the US CPI to adjust the figures to 2010 US \$. It does not make sense to use adjust for purchasing power parity here, because international donors face only nominal exchange rates.

# Results

In Figure \ref{fig:map}, I've included a map of log VELs averaged by country over an equal area projection.

```{r map, fig.cap="\\label{fig:map}Log VEL averaged by country"}
ne_countries(scale = "small", returnclass = "sf") %>%
  # equal earth projection
  st_transform(crs = "+proj=eqearth") %>%
  select(country_code = iso_a2) %>%
  left_join(
    ratios %>%
    group_by(country_code) %>%
    summarize(
      `Mean log VEL (2010 $/70 QALYs)` = 
        mean(`Mean log VEL (2010 $/70 QALYs)`)
    )
  ) %>%
  .["Mean log VEL (2010 $/70 QALYs)"] %>%
  plot
```

There are two important observation: first, that VELs differ dramatically between countries, and second, that for very many countries, there is not a single published public health cost-effectiveness study. These countries are colored in white.

I found that VELs are very strongly determined by health care spending per capita, where health care spending per capita is measured as the sum of private, public, and non-profit spending per capita. In Figure \ref{fig:plot}, I've included a scatterplot of the values.

```{r plot, fig.cap="\\label{fig:plot}VEL vs health care spending per capita"}
ggplot(together) + 
  aes(
    x = `Log health care spending per capita (2010 $)`, 
    y = `Mean log VEL (2010 $/70 QALYs)`
  ) +
  geom_point()
```

I regressed the log of VEL values on log health care spending per capita. The results are Table 1.

```{r regression}
model %>%
  pander
```

There appears to be decreasing returns to health care spending per capita: with 95% confidence, a 1% increase in health care spending per capita leads to between a `r pander(interval[[1]])`% and `r pander(interval[[2]])`% increase in the VEL.

The variance in health care spending per capita can be further decomposed by sector. In Table 2, I've included a covariance matrix for health care spending per capita, normalized so that the sum is 1.

```{r covariance_matrix}
expenditures %>%
  select(-year, -country_code) %>%
  filter(complete.cases(.)) %>%
  cov %>%
  {./sum(.)} %>%
  pander(caption = "Proportion of variation in health care spending per capita by sector")
```

The majority of differences in health care spending per capita can be explained by spending differences from the public sector, with additional components related to private health care spending per capita and the correlation between spending per capita from these two sectors. Non-profit health care spending per capita ever so slightly reduces the total variation. One might ascribe variation in public spending to nationalism; variation in private spending to disparaities in market power; and variation in external spending to discrimination. Given these interpretations, nationalism, and to a lesser extent, market power, are determining health spending priorities.

# Conclusion

The results above provide strong empirical support for the power-weighted social decision rule [@Boyce_2002]. The value of an empirical life (VEL) varies greatly between countries. For the purposes of the testing the power-weighted social decision rule, it is irrelevant whether these differences are caused by differences in private, public, or non-profit health care spending per capita. Just as important as the money that was spent saving lives is the money that was not spent saving lives. The mere existence of very cost effective ways to save lives in many countries suggests that citizens in these countries do not have the power to effectively influence global health care spending.

# Works cited